Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-26 paper.tex
Move more stuff up to the top

\subsection*{Tree sequences}
\label{sec-ts}

@@

\subsection*{Simulation interface}
\label{sec-sim-interface}

@@

\subsection*{Simulating mutations}
\label{sec-mutations}

@@

\subsection*{Selection}
\label{sec-selection}
Another ellaboration of the standard neutral coalescent with recombination
is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
In the context of the coalescent this amounts to creating a structured population
during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
in which lineages may transit between favored and unfavored backgrounds through
recombination. This allows for many selective sweep scenarios, including recurrent, partial,
and soft selective sweeps to be simulated efficiently, however in general this
efficiency comes at the cost of flexibility in comparison to forwards in
time simulation.

@@

\subsection*{Analysis of simulation results}
\label{sec-data-analysis}

\jkcomment{This is probably part of the discussion, but putting here out of
the way for now in its own subsection}

The point of running simulations is to generate data, and there
are two important ways in which the tree sequences output by
\msprime\ are more useful than classical approaches: (1) tree
sequences are more efficient; and (2) tree sequences contain
more information. We will consider these points in turn.

The classical approach to generating simulation data is
to render the results in some text format. For example,
\ms~\citep{hudson2002generating} outputs a set of sequences
and can also optionally output the marginal trees along
the genome in Newick format,
and variants of this approach are used by many simulators.
Text files have many advantages, but are slow to process
at scale. With sample sizes becoming ever larger, processing
text output can become a significant bottleneck in many analyses.
The ability to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which one can sample from the model and parameter space.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lacking a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
% JK: not sure this is worth leaving in tbh
This lack of reliable information about ancestors in simulation data
has lead, for example, to the definition of segments of
identity by descent as a set of samples having the same time
to their most recent common ancestor along some section of the
genome~\citep{yang2016rapidly}.
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

\jkcomment{Summary paragraph: tskit gives us complete information
about the genetic ancestry. and is very efficient}

% \jkcomment{A point to make somewhere in here: the tskit data model is
% integrated, with all of the information that you need in one file. Not
% splattered across a number of different files, that you have to keep track
% of.}

>>>>

% The data model
\subsection*{Tree sequences}
\label{sec-ts}

@@

% The input - how do we run simulations?
\subsection*{User interface}
\label{sec-sim-interface}

@@

% The output - what do we get out?
\subsection*{Data analysis}
\label{sec-data-analysis}
The point of running simulations is to generate data, and there
are two important ways in which the \tskit\ tree sequences output by
\msprime\ are more useful than established methods: (1) tree
sequences are more efficient; and (2) tree sequences contain
more information. We will consider these points in turn.

The standard way of representing simulation data is
to render the results in a text format. For example,
\ms~\citep{hudson2002generating} outputs a set of sequences
and can also optionally output the marginal trees along
the genome in Newick format,
and variants of this approach are used by many simulators.
Text files have many advantages, but are slow to process
at scale.
The ability to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which one can sample from the model and parameter space.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results all the more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lack a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
% JK: not sure this is worth leaving in tbh
This lack of reliable information about ancestors in simulation data
has lead, for example, to the definition of segments of
identity by descent as a set of samples having the same time
to their most recent common ancestor along some section of the
genome~\citep{yang2016rapidly}.
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

These problems, of inefficient and limited access to the simulated
data are comprehensively solved by \msprime\ and \tskit. Complete
information about ancestry and mutation simulations is immediately
available to the user in a single integrated data structure
(rather than multiple independent files), with no intermediate
processing required.
Users can subsequently compute precisely the information that they
require using the powerful suite of population genetic statistics
and other methods available in \tskit.

% Stuff we simulate: mutations
\subsection*{Simulating mutations}
\label{sec-mutations}

@@

\subsection*{Selection}
\label{sec-selection}
Another elaboration of the standard neutral coalescent with recombination
is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
In the context of the coalescent this amounts to creating a structured population
during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
in which lineages may transit between favoured and unfavoured backgrounds through
recombination. This allows for many selective sweep scenarios, including recurrent, partial,
and soft selective sweeps to be simulated efficiently, however in general this
efficiency comes at the cost of flexibility in comparison to forwards in
time simulation.

@@
